### 2010-07-21 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: uniformise notation for detagging and parenthesing
Reviewer 2 said:

"p5.  The mapping TagDesc to Desc, "de -", is replaced by "overline"
notation for indexed dataypes.  Neither notation helps me remember
what they mean - why not call it "TagDescToDesc"?"

'TagDescToDesc' is just impossible for space reason, but at least we
are now uniform in our hieroglyphism.
Desc@ICFP: tentative: put level annotation in Section 4

We shall now fulfil our implementation promises, encoding the universe
of descriptions. In and of itself, the codes, $\Desc$, is nothing but
a datatype. We are in the same situation as with $\EnumU$: we ought to
be able to describe the codes of $\Desc$ in $\Desc$ itself. Hence,
this code would be a first-class citizen, born with the standard,
generic equipment of datatypes.

@@

Our first attempt gets stuck quite quickly:
%
\[\stk{
\DescD : \Desc \\
\DescD \mapsto
  \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
       {\stkm{ \DUnit                                \\
         \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
         \DIndx{\DUnit}    }                    } \\
%\begin{array}{@{}ll}
%\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
%                               & \left[\begin{array}{l}
%                                   \DUnit                                \\
%                                   \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
%                                   \DIndx{\DUnit}                        \\
%                                 \end{array}
%                                 \right]
%\end{array}
}\]
%
Let us explain where we stand. Much as
we have done so far, we first offer a constructor choice from
$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. The reader will notice
that the `tagged' notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are defining.
For $\DUnit$, we immediately reach the end of the description. For
$\SYMBDIndx$, there is a single recursive argument. Describing
$\SYMBDSigma$ is problematic. Recall the specification of
$\SYMBDSigma$:
%
\[
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
\]
%
So, we first pack a  $\Set$, $\V{S}$. We should then like
a recursive argument \emph{indexed} by $\V{S}$, but
that is an \emph{exponential}, and our presentation is entirely
first-order so far, delivering only sums-of-products. To code our
universe, we must first enlarge it!

@@

In order to capture a notion of higher-order induction, we add a code
$\SYMBDHindx$ that takes an indexing set $\V{H}$. Intuitively,
$\SYMBDHindx$ gives a recursive subobject for each element of $H$.
%
\[\stk{
    \DHindx{(\Bhab{\M{H}}{\Set})}{(\Bhab{\M{D}}{\Desc})} : \Set \smallskip \\
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     \mapsto \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
}\]

@@

We are now able to describe our universe of datatypes:
%
\[\stk{
\DescD : \Desc \\
\DescD \mapsto\tada
{\stkm{ \DUnit\\ \SYMBDSigma\\ \SYMBDIndx\\ \SYMBDHindx}}
{\stkm{
\DUnit                                            \\
                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
}}
}\]
%
The $\DUnit$ and $\SYMBDIndx$ cases remain unchanged, as expected. We
successfully describe the $\SYMBDSigma$ case, by a simple appeal to
the higher-order induction on $\V{S}$. The $\SYMBDHindx$ case
consists in packing a $\Set$ with a recursive argument.

At a first glance, we have achieved our goal. We have described the
codes of the universe of descriptions. Taking the fixpoint of \(\descop{\DescD}\)
gives us a datatype exactly like $\Desc$. Might we be
so bold as to take \(\Desc \mapsto \Mu{\DescD}\) as the levitating
definition? If we do, we shall come down with a bump! 
To complete our levitation, just as in the magic trick, requires
hidden assistance. Let us explain the problem and reveal the `invisible
cable' which fixes it.

@@

The definition \(\Desc \mapsto \Mu{\DescD}\) is circular,
but the offensive recursion is concealed by a prestidigitation.
Expanding \(\toDesc{-}\) and propagating types as in
Figure~\ref{fig:type-checking} reveals the
awful truth:
\[
\Desc\mapsto
\Mu{\stk{(\DSigma{\EnumT{\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}}\\}
    {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:(\LAM{\_}\Desc)\\
\;\sqr{\stkm{
\DUnit                                            \\
                                   \DSigma{\Set}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
                                   \DIndx{\DUnit}                                    \\
                                   \DSigma{\Set}{\LAM{\_} \DIndx{\DUnit}}
}}})}}
\]
The recursion shows up only because we must specify the return type
of the general-purpose \(\F{switch}\), and it is computing a \(\Desc\)!
Although type propagation
allows us to hide this detail \emph{when defining a function}, we cannot
readily suppress this information and check types
when \(\F{switch}\) is fully applied.

@@

We are too close to give up now. If only we did not need to
supply that return type, especially when we know what it must be.
We eliminate the recursion by \emph{specialising} \(\F{switch}\):
%
\[
\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
                \EnumT{\V{E}} \To \Desc
\]
%
The magician's art rests here, in this extension. We conceal it
behind a type propagation rule for \(\F{switchD}\) which we apply
with higher priority than for \(\F{switch}\) in general.
%
\[
\Rule{\Gamma \Vdash
  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{(\PLAM{x}{\EnumT{E}}\Desc)}}}
                           {\M{t'}}}
     {\Gamma \Vdash
\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
 {\switchD{\M{E}}{\M{t'}}}}
\]
As a consequence, our definition above now propagates without
introducing recursion. Of course, by pasting together the declaration
of \(\Desc\) and its internal copy, we have made it appear in its own
type. Hardwired as a \emph{fait accompli}, this creates no regress,
although one must assume the definition to recheck it.\footnote{Our
  Agda model does not formalize the $\SYMBswitchD$
  construction. However, it proves the isomorphism between the defined
  descriptions and the coded ones. The role of $\SYMBswitchD$ is to
  collapse this isomorphism, hence identifiying defined and coded
  descriptions.}

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a readily propagated non-dependent return type $\V{T}$.
Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
}\]
%
We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
shape, but its sub-elements are of the wrong type. On the other hand,
for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
element in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
\(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
program to do that---please forgive us if we lapse to a pattern matching
notation, for readability's sake.
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}             \mapsto
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} \mapsto        \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                      \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
}\]
%
Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
type theory, we have built a generic catamorphism. Any datatype will
now come equipped with this operation, for free.

@@

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as many
arguments to $\cata$ as possible:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})}) \To \\
         & \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{X}})})} \To
           \Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{(\FreeMonad{\V{D}}{\V{X}})})\: 
          (\Mu{(\toDesc{(\FreeMonad{\V{D}}{\V{Y}})})})\: 
          \SHED
}\]
%
We are left with implementing the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
         & \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]

>>>>

We shall now fulfil our implementation promises, encoding the universe
of descriptions. In and of itself, the codes, $\Desc_n$, is nothing
but a datatype in $\Set_{n+1}$. We are in the same situation as with
$\EnumU$: we ought to be able to describe the codes of $\Desc_n$ in
$\Desc_{n+1}}$ itself. Hence, this code would be a first-class
  citizen, born with the standard, generic equipment of datatypes.

@@

Our first attempt gets stuck quite quickly:
%
\[\stk{
\DescD_n : \Desc_{n+1} \\
\DescD_n \mapsto
  \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
       {\stkm{ \DUnit                                \\
         \DSigma{\Set_{n}}{(\LAM{\V{S}} \SHED)}      \\
         \DIndx{\DUnit}    }                    } \\
%\begin{array}{@{}ll}
%\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
%                               & \left[\begin{array}{l}
%                                   \DUnit                                \\
%                                   \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
%                                   \DIndx{\DUnit}                        \\
%                                 \end{array}
%                                 \right]
%\end{array}
}\]
%
Let us explain where we stand. Much as
we have done so far, we first offer a constructor choice from
$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. The reader will notice
that the `tagged' notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are defining.
For $\DUnit$, we immediately reach the end of the description. For
$\SYMBDIndx$, there is a single recursive argument. Describing
$\SYMBDSigma$ is problematic. Recall the specification of
$\SYMBDSigma$:
%
\[
\DSigma{(\Bhab{\V{S}}{\Set_n})}{(\Bhab{\V{D}}{\V{S} \To \Desc_n})} : \Desc_n\\
\]
%
So, we first pack a  $\Set$, $\V{S}$. We should then like
a recursive argument \emph{indexed} by $\V{S}$, but
that is an \emph{exponential}, and our presentation is entirely
first-order so far, delivering only sums-of-products. To code our
universe, we must first enlarge it!

@@

In order to capture a notion of higher-order induction, we add a code
$\SYMBDHindx$ that takes an indexing set $\V{H}$. Intuitively,
$\SYMBDHindx$ gives a recursive subobject for each element of $H$.
%
\[\stk{
    \Bhab{\DHindx{(\Bhab{\M{H}}{\Set_n})}
                 {(\Bhab{\M{D}}{\Desc_n})}}
         {\Desc_n}
\smallskip \\
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}} 
    \mapsto 
        \TIMES{(\V{H} \To \V{X})}
              {\descop{\V{D}}{\V{X}}} \\
}\]

@@

We are now able to describe our universe of datatypes:
%
\[\stk{
\DescD_n : \Desc_{n+1} \\
\DescD_n \mapsto\tada
{\stkm{ \DUnit\\ \SYMBDSigma\\ \SYMBDIndx\\ \SYMBDHindx}}
{\stkm{ \DUnit  \\
        \DSigma{\Set_n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}}   \\
        \DIndx{\DUnit}                                    \\
        \DSigma{\Set_n}{\LAM{\_} \DIndx{\DUnit}}
}}
}\]
%
The $\DUnit$ and $\SYMBDIndx$ cases remain unchanged, as expected. We
successfully describe the $\SYMBDSigma$ case, by a simple appeal to
the higher-order induction on $\V{S}$. The $\SYMBDHindx$ case
consists in packing a $\Set$ with a recursive argument.

At a first glance, we have achieved our goal. We have described the
codes of the universe of descriptions. Taking the fixpoint of \(\descop{\DescD_n}\)
gives us a datatype exactly like $\Desc_n$. Might we be
so bold as to take \(\Desc_n \mapsto \Mu{\DescD_n}\) as the levitating
definition? If we do, we shall come down with a bump! 
To complete our levitation, just as in the magic trick, requires
hidden assistance. Let us explain the problem and reveal the `invisible
cable' which fixes it.

@@

The definition \(\Desc_n \mapsto \Mu{\DescD_n}\) is circular,
but the offensive recursion is concealed by a prestidigitation.
Expanding \(\toDesc{-}\) and propagating types as in
Figure~\ref{fig:type-checking} reveals the
awful truth:
\[
\Desc_n \mapsto
    \Mu{\stk{(
        \DSigma{\EnumT{\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}}\\}
               {\;\SYMBswitch\:\sqr{ \DUnit\:\: \SYMBDSigma\:\: \SYMBDIndx\:\: \SYMBDHindx}\:
                                   (\LAM{\_}\Desc)\\
                             \;\sqr{\stkm{
                                   \DUnit \\
                                   \DSigma{\Set_n}{\LAM{\V{S}} \DHindx{\V{S}}{\DUnit}} \\
                                   \DIndx{\DUnit} \\
                                   \DSigma{\Set_n}{\LAM{\_} \DIndx{\DUnit}}}}})}}
\]
%%
The recursion shows up only because we must specify the return type
of the general-purpose \(\F{switch}\), and it is computing a \(\Desc\)!
Although type propagation
allows us to hide this detail \emph{when defining a function}, we cannot
readily suppress this information and check types
when \(\F{switch}\) is fully applied.

@@

We are too close to give up now. If only we did not need to
supply that return type, especially when we know what it must be.
We eliminate the recursion by \emph{specialising} \(\F{switch}\):
%
\[
\F{switchD} : \PITEL{\V{E}}{\EnumU}   \To
                (\spi{\V{E}}{\LAM{\_} \Desc}) \To
                \EnumT{\V{E}} \To \Desc
\]
%
The magician's art rests here, in this extension. We conceal it
behind a type propagation rule for \(\F{switchD}\) which we apply
with higher priority than for \(\F{switch}\) in general.
%
\[
\Rule{\Gamma \Vdash
  \propag{\push{\sqr{\vec{t}}}{\spi{\M{E}}{\PLAM{x}{\EnumT{E}}\Desc}}}
                           {\M{t'}}}
     {\Gamma \Vdash
\propag{\push{\sqr{\vec{t}}}{\EnumT{\M{E}}\To\Desc}}
 {\switchD{\M{E}}{\M{t'}}}}
\]
As a consequence, our definition above now propagates without
introducing recursion. Of course, by pasting together the declaration
of \(\Desc\) and its internal copy, we have made it appear in its own
type. Hardwired as a \emph{fait accompli}, this creates no regress,
although one must assume the definition to recheck it.\footnote{Our
  Agda model does not formalize the $\SYMBswitchD$
  construction. However, it proves the isomorphism between the defined
  descriptions and the coded ones. The role of $\SYMBswitchD$ is to
  collapse this isomorphism, hence identifiying defined and coded
  descriptions.}

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a readily propagated non-dependent return type $\V{T}$.
Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
}\]
%
We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
shape, but its sub-elements are of the wrong type. On the other hand,
for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
element in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
\(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
program to do that---please forgive us if we lapse to a pattern matching
notation, for readability's sake.
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}             \mapsto
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} \mapsto        \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}                                      \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
}\]
%
Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
type theory, we have built a generic catamorphism. Any datatype will
now come equipped with this operation, for free.

@@

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as many
arguments to $\cata$ as possible:
%
\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
         & \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{X}}})} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})} 
\end{array} \\
\subst\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma} \mapsto
  \cata\: (\toDesc{\FreeMonad{\V{D}}{\V{X}}})\: 
          \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}\: 
          \SHED
}\]
%
We are left with implementing the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:
%
\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} 
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}) \To \\
         & \descop{\toDesc{\FreeMonad{\V{D}}{\V{X}}}}{\Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}} \To
           \Mu{(\toDesc{\FreeMonad{\V{D}}{\V{Y}}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{x}                   \\
\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}\: \pair{\V{c}}{\V{xs}}{} & \Con{\pair{\V{c}}{\V{xs}}{}}
\end{array}
}\]
